\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-24 paper.qmd
Add three parameter choices can discuss with star data set

```{r}
#| echo: false
#| fig-cap: Goodness of fit statistics from different n_neighbors parameter of UMAP applied to training S-curve dataset. What is the best parameter choice in UMAP to represent the original data in 2D?
#| label: fig-diagnosticpltDiffParam
#| out-width: 100%

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/x5VPB-wOLm4>), (b) Model in the 2D space with UMAP (<https://youtu.be/Cm1dW6iLyHQ>), (c) Model in the 2D space with PHATE (<https://youtu.be/HB0Y3ilz6sI>), (d) Model in the 2D space with TriMAP (<https://youtu.be/iSVkXIrOrSA>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/O2QBctsc4uM>).
#| out-width: 100%

@@

```{r}
#| echo: false
#| fig-cap: Goodness of fit statistics from different NLDR techniques applied to training five spherical Gaussian cluster dataset. What is the best NLDR technique to represent the original data in 2D?
#| label: fig-diagnosticpltGau
#| out-width: 100%

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisPBMCUMAP
 #| out-width: 100%

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.1) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/SO3thJ8PMi8>)"
#| label: fig-nldervisPBMCUMAP2
 #| out-width: 100%

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.01) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/hP8ie_oMXw0>)"
#| label: fig-nldervisPBMCUMAP3
 #| out-width: 100%

@@

- In 2D, there are two well-separated clusters

@@

- In high-D view, the clusters having some non-linearity patter, which is also not captured in tSNE. Therefore, this is not a good choice.

@@

@@

plot_list2_harps <- tSNE_data_harps |>
    ggplot(aes(x = tSNE1,
               y = tSNE2, color = cluster))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#cab2d6", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  guides(color = guide_legend(title = "cluster"),
         color = guide_legend(nrow = 3)) +
 annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

trimesh_removed_tsne_harps <- trimesh_removed_tsne_harps +
  # xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  # theme(axis.text = element_text(size = 5),
  #       axis.title = element_text(size = 7)) +
  geom_point(colour = "#33a02c", size = 0.5) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from tSNE (perplexity = 40) applied for the HARPS GTO dataset. Is this a best representation of the original data?, (b) Model in the 2D space with tSNE (<https://youtu.be/txTD7GTxt6Y>)"
#| label: fig-nldervisharpstSNE

>>>>

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from different n_neighbors parameter of UMAP applied to training S-curve dataset. What is the best parameter choice in UMAP to represent the original data in 2D?"
#| label: fig-diagnosticpltDiffParam
#| out-width: 100%

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: "Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/x5VPB-wOLm4>), (b) Model in the 2D space with UMAP (<https://youtu.be/Cm1dW6iLyHQ>), (c) Model in the 2D space with PHATE (<https://youtu.be/HB0Y3ilz6sI>), (d) Model in the 2D space with TriMAP (<https://youtu.be/iSVkXIrOrSA>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/O2QBctsc4uM>)."
#| out-width: 100%

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from different NLDR techniques applied to training five spherical Gaussian cluster dataset. What is the best NLDR technique to represent the original data in 2D?"
#| label: fig-diagnosticpltGau
#| out-width: 100%

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisPBMCUMAP
##| out-width: 100%
#| fig-pos: h

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.1) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/SO3thJ8PMi8>)"
#| label: fig-nldervisPBMCUMAP2
##| out-width: 100%
#| fig-pos: h

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.01) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/hP8ie_oMXw0>)"
#| label: fig-nldervisPBMCUMAP3
##| out-width: 100%
#| fig-pos: h

@@

- In 2D, there are two well-separated clusters, but the clusters are really close

@@

- In high-D view, the clusters having some non-linearity patter, which is also not captured in tSNE. Therefore, this is not a good choice. Also, the two clusters are really near to each other and can't see any separation. Also continuity of the structure. Long edges are presence also.

@@

<!-- Author suggestion-->
```{r}
#| warning: false
#| echo: false

@@

plot_list2_harps <- tSNE_data_harps |>
    ggplot(aes(x = tSNE1,
               y = tSNE2, color = cluster))+
    geom_point(alpha=0.5) +
    coord_cartesian(xlim = c(-25, 25), ylim = c(-12, 12)) +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#cab2d6", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  guides(color = guide_legend(title = "cluster"),
         color = guide_legend(nrow = 3)) +
 annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

trimesh_removed_tsne_harps <- trimesh_removed_tsne_harps +
  # xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  # theme(axis.text = element_text(size = 5),
  #       axis.title = element_text(size = 7)) +
  geom_point(colour = "#33a02c", size = 0.5) +
  #coord_cartesian(xlim = c(-25, 25), ylim = c(-12, 12)) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from tSNE (perplexity = 40) applied for the HARPS GTO dataset. Is this a best representation of the original data?, (b) Model in the 2D space with tSNE (<https://youtu.be/jxo_rPyGv9k>)"
#| label: fig-nldervisharpstSNE
#| fig-pos: h

<!-- param set 1-->

- 5 clusters shows in 2D

- 3 clusters which are really close shows in high_d data with the model

tSNE_data_harps <- read_rds("data/harps/harps_tsne_23.rds")

plot_list2_harps <- tSNE_data_harps |>
    ggplot(aes(x = tSNE1,
               y = tSNE2, color = clusters))+
    geom_point(alpha=0.5) +
    coord_cartesian(xlim = c(-22, 18), ylim = c(-27, 32)) +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#cab2d6", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  guides(color = guide_legend(title = "cluster"),
         color = guide_legend(nrow = 3)) +
 annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

## Import data
# training_data_1 <- read_rds("data/harps/harps_data.rds")
# training_data_1 <- training_data_1 |>
#   mutate(ID = 1:NROW(training_data_1))
# 
# clusters <- training_data_1$cluster
# select_names <-  names(training_data_1)[-c(18, length(training_data_1))]
# 
# training_data_1 <- training_data_1 |> select(-cluster)
# names(training_data_1) <- append(paste0("x", 1:17), "ID")
# 
# tSNE_data_harps <- tSNE_data_harps |>
#   select(tSNE1, tSNE2) |>
#   mutate(ID = 1:NROW(tSNE_data_harps))

num_bins_tsne_harps <- calculate_effective_x_bins(.data = tSNE_data_harps, x = tSNE1,
                                                cell_area = 1) #43
num_bins_tsne_harps <- 30

trimesh_removed_tsne_harps <- trimesh_removed_tsne_harps +
  # xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  # theme(axis.text = element_text(size = 5),
  #       axis.title = element_text(size = 7)) +
  geom_point(colour = "#33a02c", size = 0.5) +
  #coord_cartesian(xlim = c(-22, 18), ylim = c(-27, 32)) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from tSNE (perplexity = 23) applied for the HARPS GTO dataset. Is this a best representation of the original data?, (b) Model in the 2D space with tSNE (<https://youtu.be/_is9qStggFE>)"
#| label: fig-nldervisharpstSNE2
#| fig-pos: h

<!-- param set 2-->

tSNE_data_harps <- read_rds("data/harps/harps_tsne_91.rds")

plot_list2_harps <- tSNE_data_harps |>
    ggplot(aes(x = tSNE1,
               y = tSNE2, color = clusters))+
    geom_point(alpha=0.5) +
    coord_cartesian(xlim = c(-8, 7), ylim = c(-10, 16)) +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#b2df8a", "#33a02c", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#cab2d6", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  guides(color = guide_legend(title = "cluster"),
         color = guide_legend(nrow = 3)) +
 annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

## Import data
# training_data_1 <- read_rds("data/harps/harps_data.rds")
# training_data_1 <- training_data_1 |>
#   mutate(ID = 1:NROW(training_data_1))
# 
# clusters <- training_data_1$cluster
# select_names <-  names(training_data_1)[-c(18, length(training_data_1))]
# 
# training_data_1 <- training_data_1 |> select(-cluster)
# names(training_data_1) <- append(paste0("x", 1:17), "ID")
# 
# tSNE_data_harps <- tSNE_data_harps |>
#   select(tSNE1, tSNE2) |>
#   mutate(ID = 1:NROW(tSNE_data_harps))

num_bins_tsne_harps <- calculate_effective_x_bins(.data = tSNE_data_harps, x = tSNE1,
                                                cell_area = 1) #43
#num_bins_tsne_harps <- 30

trimesh_removed_tsne_harps <- trimesh_removed_tsne_harps +
  # xlab(expression(C[x]^{(2)})) + ylab(expression(C[y]^{(2)})) +
  # theme(axis.text = element_text(size = 5),
  #       axis.title = element_text(size = 7)) +
  geom_point(colour = "#33a02c", size = 0.5) +
  #coord_cartesian(xlim = c(-8, 7), ylim = c(-10, 16)) +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from tSNE (perplexity = 91) applied for the HARPS GTO dataset. Is this a best representation of the original data?, (b) Model in the 2D space with tSNE (<https://youtu.be/aGvHFhWw_Xs>)"
#| label: fig-nldervisharpstSNE3
#| fig-pos: h

### 2024-01-24 paper.tex
Add three parameter choices can discuss with star data set

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  In 2D, there are two well-separated clusters
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
\item
  The parameter is, in a sense, a guess about the number of close
  neighbors each point has.
\end{itemize}

@@

\caption{\label{fig-nldervisharpstSNE}(a) 2D layout from tSNE
(perplexity = 40) applied for the HARPS GTO dataset. Is this a best
representation of the original data?, (b) Model in the 2D space with
tSNE (\url{https://youtu.be/txTD7GTxt6Y})}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, frame hidden, interior hidden, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  In 2D, there are two well-separated clusters, but the clusters are
  really close
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
  Also, the two clusters are really near to each other and can't see any
  separation. Also continuity of the structure. Long edges are presence
  also.
\item
  The parameter is, in a sense, a guess about the number of close
  neighbors each point has.
\end{itemize}

\begin{figure}[h]

@@

\caption{\label{fig-nldervisharpstSNE}(a) 2D layout from tSNE
(perplexity = 40) applied for the HARPS GTO dataset. Is this a best
representation of the original data?, (b) Model in the 2D space with
tSNE (\url{https://youtu.be/jxo_rPyGv9k})}

\begin{itemize}
\item
  5 clusters shows in 2D
\item
  3 clusters which are really close shows in high\_d data with the model
\end{itemize}

\begin{figure}[h]

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisharpstSNE2-1.pdf}

\caption{\label{fig-nldervisharpstSNE2}(a) 2D layout from tSNE
(perplexity = 23) applied for the HARPS GTO dataset. Is this a best
representation of the original data?, (b) Model in the 2D space with
tSNE (\url{https://youtu.be/_is9qStggFE})}

\begin{figure}[h]

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisharpstSNE3-1.pdf}

\caption{\label{fig-nldervisharpstSNE3}(a) 2D layout from tSNE
(perplexity = 91) applied for the HARPS GTO dataset. Is this a best
representation of the original data?, (b) Model in the 2D space with
tSNE (\url{https://youtu.be/aGvHFhWw_Xs})}
